Demographic consequences of partner diversity and turnover in a multi-species ant-plant mutualism

### 2025-06-18 Manuscript/ant_cactus_SUBMISSION2.tex
Manuscript Updates

	\section*{Abstract}
% 200 words max
Mutualisms commonly involve multiple partner species but the ecological consequences of partner diversity remain poorly understood.
Partner diversity may benefit a focal mutualist through mechanisms that mirror positive effects of species diversity on ecosystem function, such as complementarity or portfolio effect. 
We studied the fitness consequences of partner diversity and turnover in the multi-species food-for-protection mutualism involving an EFN-bearing plant (tree cholla cactus \textit{Cylindropuntia imbricata}) and multiple species of nectar-feeding ants (\textit{Crematogaster opuntiae}, \textit{Liometopum apiculatum}, and other less frequent ants) that provide defense from herbivores. 
We used long-term data and Bayesian hierarchical statistical models to determine the demographic impacts of different ant partners, and the dynamics of turnover from one exclusive partner association to another. 
We then constructed a stochastic, multi-state Integral Projection Model (IPM) with which we simulated tree cholla populations with different richness and composition of partners.
We found that, while ant partners had different impacts on plant vital rates (\textit{C. opuntiae}-tended plants had advantages in growth and survival when small, and \textit{L. apiculatum}-tended plants had floral viability advantages), plant fitness was effectively insensitive to partner identity and number, because effects on the highest-sensitivity vital rates were consistent across partner species. 
Furthermore, there was no evidence that benefits of partner diversity manifest in the context of fluctuating environments (i.e., the portfolio effect). 
In contrast to much of the previous literature, this study highlights that demographic benefits of mutualism can be surprisingly robust to diversity and turnover of partners.

\newpage{}
\section*{Introduction}
Mutualisms are species interactions where all participants receive net benefits, leading to higher individual fitness and increased population growth rates. 
They are widespread species interactions but can deteriorate into commensalism or parasitism under conditions that elevate costs or dampen benefits \citep{Rodriguez-Rodriguez2017,Song2020,Mandyam2014,Thrall2007, Bahia2022,Bronstein1994,Chamberlain2014,Frederickson2013,Axelrod1981,Leigh2010}.
Mutualisms are considered more context dependent than other species interactions \citep{Chamberlain2014,Frederickson2013}, meaning the magnitude and sign of interaction strength are often determined by environmental conditions and species' identities \citep{Noe1994,Leigh2010}.

Mutualism is defined at the level of a species pair (+/+) but these interactions are embedded within multi-species communities, and growing evidence suggests that pairwise interactions are poor predictors of the net effects of multi-species mutualism \citep{Afkhami2014,Palmer2010,Bascompte2009,Dattilo2014}. 
A focal mutualist may interact with multiple guilds of partner types (e.g., plants that interact with pollinators, seed dispersers, soil microbes, and ant defenders) or with multiple partner species within the same guild (e.g., plants visited by multiple pollinator species). 
Within a mutualist guild, partner species often differ in the amount or type of goods or services they provide, making partner identity an important source of contingency in mutualism \citep{Stanton2003}. 
Whether and how partner diversity modifies the demographic effects of mutualistic interactions remain open questions within relevance in applied settings \citep{rogers2014, Thibaut2012, Frederickson2005, Palmer2010}.

There are multiple mechanisms by which partner diversity can influence the net benefits accrued by a focal mutualist, mirroring the mechanisms by which, at a larger scale of organization, biodiversity can influence ecosystem function \citep{Barrett2015,Ushio2020,Winfree2020}.
First, when there is a hierarchy of fitness effects (a consistent ranking of best to worst mutualists), a more diverse sample of the partner community may be more likely to include the best partner \citep{Frederickson2013}.
This can lead to an apparent benefit of diversity driven by a sampling effect \citep{Batstone2018} -- but there is no benefit of diversity \emph{per se}, only better and worse partners. 
However, if partner associations are mutually exclusive (a focal mutualist may engage with only one partner at a time), then partner diversity may impose opportunity costs, leading to negative effects of a diverse mutualist assemblage relative to exclusive association with the single best partner \citep{Miller2007}. 
Second, even within a single mutualist guild, the benefits conferred by alternative partner species can vary in type and not just degree \citep{Stachowicz2005,Bronstein2006,Stanton2003}. 
This can lead to a positive effect of partner diversity through complementarity of alternative functions \citep{Batstone2018}. 
Positive diversity effects through complementarity need not be additive: interference or synergies between partners can make their combined effect different than the expected from the sum of complementary functions \citep{Afkhami2014}. 
Third, partner species can have species-specific responses to environmental variation, either spatially \citep{Ollerton2006} or temporally \citep{Alarcon2008}. 
Multiple partners can therefore act as a 'portfolio' that stabilizes fitness benefits across spatial or temporal heterogeneity, leading to positive effects of partner diversity through the portfolio effect \citep{Batstone2018,Lazaro2022,Horvitz1990}.

Partner diversity can have different effects on a focal mutualist depending on whether partners are present simultaneously or sequentially (partner turnover) \citep{Djieto-Lordon2005, Ness2006, Bruna2014,Barrett2015,Ushio2020,Dattilo2014}. 
Sequential associations are likely when alternative partners engage in interference competition for access to a shared mutualist \citep{Kiers2003,Batstone2018,Tgaard2015,Wulff2008}. 
Turnover can happen at different timescales, from minutes to years \citep{Oliveira1999,Horvitz1986}. 
The frequency of partner turnover can impact the level of benefits received by the focal mutualist, particularly if the benefits continue to accumulate with successive turnover (e.g., when sequential partners provide complementary functions) or if they saturate over time \citep{Sachs2004,Fiala1994}.
Directionality of turnover can also influence effects of partner diversity if partner identity changes consistently across ontogeny of a focal mutualist \citep{Fonseca2003,Noe1994,Dejean2008}.
For example, plant susceptibility to enemies can change across life stages \citep{Boege2005,Barton2010}, so the benefits of a diverse guild of defensive mutualists are greatest when more defensive partner species align with more vulnerable life stages \citep{Djieto-Lordon2005,Dejean2008}.
Thus, the duration and order of interactions may be just as important as the identities of the interacting partners.

Defensive ant-plant mutualisms -- where plants provide food and/or housing to ants that in turn defend them from enemies -- are widespread interactions that offer valuable model systems for the ecology and evolution of mutualism \citep{Bronstein1998, Bronstein2006}. 
Extrafloral nectar (EFN) -bearing plants can serve as dietary resources that promote ant abundance and colony size \citep{Byk2011, Ness2009, Ness2006, Donald2022}.
In turn, presence of defensive ant partners is often linked to reductions in herbivory  \citep{Trager2010, Rudgers2004} and demographic advantages for the plant partner \citep{Baez2016}.
Defensive ant-plant mutualisms are commonly multi-species, where a guild of ant partner species share, and often compete for, a plant mutualist \citep{Bronstein1998, Beattie1985, Trager2010, Agrawal1998}.
Ant partners can vary in their ability to deter herbivores \citep{Bruna2014}, and visitation by low quality ant partners can prevent visitation by higher quality partners, causing a reduction in fitness through missed opportunity costs \citep{Fraser2001, Frederickson2005}.
Susceptibility to herbivory can also vary significantly throughout the life stages of the plant \citep{Boege2005}, suggesting that the order and timing of successive partners is important to the fitness impacts of the combined partner guild \citep{Barton2010, Boege2005, Fonseca2003}.
Finally, herbivore identity and pressure can vary inter-annually \citep{Wetzel2023}, much like mutualist identity and presence, meaning the threat plants face can vary just as much as the protection they receive due to temporal stochasticity. 
Previous studies have investigated how ant partner diversity affects plant fitness \citep{Palmer2010,Fiala1994,Gaume1998,Dattilo2014,Ludka2015}. 
However, little is known about the combined effects of partner identity, directional partner turnover, and temporal stochasticity, particularly because the long-term data needed to quantify fluctuations in ant-plant interactions are rarely available.

This study examined the consequences of partner diversity in a food-for-protection mutualism between the tree cholla cactus (\textit{Cylindropuntia imbricata}), a long-lived EFN-bearing plant, and multiple species of ant partners.
Tree cholla are tended by two common ant partners and several additional rarer species, all of which collect EFN in exchange for defense against herbivores.
These ant species locally co-occur but individual plants are typically tended by only one species that patrols the plant around-the-clock and maintains control of the plant's nectar resources, usually for an entire growing season \citep{Ohm2014, Donald2022}. 
Switches between partner species, or between vacancy and ant occupancy, commonly occur from one growing season to the next \citep{Miller2007}. 
Prior experiments suggested a hierarchy of mutualist quality, with \textit{L. apiculatum} providing strong anti-herbivore defense and \textit{C. opuntiae} having net negative effects because herbivore deterrence is outweighed by deterrence of pollinators \citep{Miller2007,Ohm2014}. 
However, previous studies in this system focused on single life stages (adult plants) or vital rates (seed production), did not integrate the demographic effects of ant defense across the life cycle, and did not incorporate temporal stochasticity, all of which may be essential for understanding net fitness effects of a diverse partner guild in variable environments.

Here we used a unique long-term data set that allows us to explore mutualistic associations with multiple partner species, longitudinal dynamics of partner turnover, and demographic effects of alternative partners that vary across plant size structure and across nearly 20 years of inter-annual fluctuations. 
We used this observational data set of plant demography and ant-plant associations, contextualized by previous ant exclusion experiments, to investigate whether and through which mechanism(s) partner diversity affects the fitness benefits of ant visitation. 
Specifically, we asked:
	\begin{enumerate}	
		\item{What are the vital rate effects of association with alternative partners and how do these effects fluctuate across years?}
		\item{What are the frequency and direction of partner turnover across the plant life cycle?}	
		\item{What is the net effect of partner diversity on plant fitness, and what mechanism(s) explain(s) this effect?}
	\end{enumerate}
To answer these questions we used a hierarchical Bayesian statistical approach to estimate demographic vital rates for hosts in different states of ant occupancy and to quantify state-dependent partner turnover from the long-term data. 
We then used a stochastic, multi-state integral projection model (IPM) that combines diverse effects on vital rates and pathways of partner turnover to quantify effects of partner diversity on plant fitness.

@@

\subsection*{Data Collection}
This study is based on long-term data spanning 2004 to 2023 that documented plant demographic rates in relation to their ant partner status. 
From 2004 to 2008, we censused 134 plants distributed across three spatial blocks. 
This initial census group was discontinued in 2009, when we established six 30 $\times$ 30-meter plots and tagged all tree cholla within those plots. 
Two additional 30 $\times$ 30-meter plots were added in 2011, and this group of eight plots (mean $\pm$ SD Tree cholla per plot: 87.30 $\pm$ 18.28) has since been censused annually through 2023 (with the exception of 2020 due to the pandemic shutdown). 
For all plants, in May or early June of each year we recorded plant survival since the last survey and, for survivors, we recorded height (cm), maximum crown width (cm), and crown width perpendicular to the maximum (cm).
Size measurements were used to calculate plant volume ($cm^3$) based on the volume of an elliptical cone. 
We measured reproduction by counting flowerbuds, and in most years we distinguished between flowerbuds that were viable versus aborted. 
We recorded the ant species presence and identity (or vacancy if no ants present).
Occurrences of more than one ant species on one plant were rare (less than 5\% of observations), and for the purpose of this analysis we classified the plant as being occupied by the more abundant species. 
Plots were searched for new recruits each year, and these were added to the census.
These data allowed us to link each plant's demographic fate (survival, growth, and reproduction) to its state of ant visitation. 
In total, the data set includes a total of 9,787 observations of 1,141 individuals across 15 complete transition years (spanning May/June of year $t$ to May/June of year $t+1$) \citep{DataCholla}. 
In addition to missing the year 2020, there are gaps in the time series where we switched plots or plants (and thus broke up transition years for growth/survival and partner turnover) or where we did not distinguish between viable and aborted flower buds.

To complete the tree cholla life cycle, we used additional, smaller data sets from previously published studies to estimate seed and seed bank parameters. 
Ohm et al. \citeyear{Ohm2014} provide data on the number of seeds per fruit for plants tended by \textit{L. apiculatum}, \textit{C. opuntiae}, or no ants (experimental exclusion), accounting for their effects on pollinator visitation. 
Elderd and Miller \citeyear{elderd2016quantifying} provide data on seed entry to the seed bank and seedling germination and survival rates.

@@

\subsubsection*{Partner diversity simulation experiments}
Using the fully parameterized multi-state IPM, we conducted simulation experiments to quantify how diversity and identity of ant partners influenced plant fitness. 
From the full version of the model (described above) corresponding to the observed assemblage of partners and observed frequencies of partner transition, we created treatments corresponding to all eight ``counter-factual'' scenarios of diversity and composition: no ant partners (complete vacancy); one ant partner (\textit{C. opuntiae} only, \textit{L. apiculatum} only, Other only); two partners (all pairwise combinations of \textit{C. opuntiae}, \textit{L. apiculatum}, and Other); and three partners (observed scenario of all ant states).
These simulation experiments were made possible by extrapolating ant-specific demographic performance across the size distribution, even for combinations of size and ant occupancy that were rarely observed. 
For example, the no-partner scenario modeled a hypothetically ant-free cactus population, even though no such population exists to our knowledge, by applying the statistical knowledge gleaned from vacant plants across the size distribution. 
We refer to stochastic fitness associated with partner number or identity using superscripts, e.g. $\lambda^{0}_{S}$ for vacant plants (zero partners), $\lambda^{1+}_{S}$ for any state of ant tending, $\lambda^{C}_{S}$ for tending by only \textit{C. opuntiae}, $\lambda^{CO}_{S}$ for tending by \textit{C. opuntiae} and Other ants, etc.

In all scenarios that included any ant partners, we preserved the observed pattern of size-dependent vacancy/occupancy (estimated through the ant transition sub-model) and manipulated partner identity conditional on occupancy by any ant. 
This means, for example, that the \textit{C. opuntiae}-only scenario included two possible states, vacancy and occupied by \textit{C. opuntiae}. 
While our statistical models allow us to extrapolate the demographic performance of ant-tended plants to small sizes that are typically vacant, the natural history of this system tells us that this is not biologically sensible. 
Small, non-reproductive plants are typically vacant because they do not produce extrafloral nectar, and once plants begin producing nectar they are nearly always ant-tended \citep{Miller2014}. 
Our simulation experiments preserved this basic biology, avoiding tiny ant-occupied plants that do not and could not occur in nature.

@@

\begin{figure}[H]
	\includegraphics[width = 0.75\linewidth]{Figures/grow_contour_v2.png}
	\caption{The next predicted size of tree cholla based on previous size in relation to ant partner (a, \textit{C. opuntiae}; b, \textit{L. apiculatum}; c, Other; d, Vacant, e, all states combined). Points show observed size transitions, colored lines are the mean next size, and contours show probability density of the skewed normal growth model.  
	Dashed lines indicate extrapolation while solid lines indicate the range of observed data.}
	\label{fig:Grow}
\end{figure}

@@

\begin{figure}[H]
	\includegraphics[width=0.95\linewidth]{Figures/transition.png}
	\caption{The probability of ant state based on size and previous ant state. Each panel shows size-dependent probabilities of the next ant state conditional on the previous ant state and size. Solid lines represent predictions of the multinomial statistical model and points show data binned over size intervals, where point size is proportional to the number of observations.}
	\label{fig:Ant_Transition}
\end{figure}

>>>>

	\section*{Abstract} \label{res:abstract}
% 200 words max
The ecological consequences of sequential multi-species mutualisms, which involve multiple partners across time, with regards to partner diversity and turnover remain poorly understood.
In such sequential mutualisms, partner identity can change over time due to environmental variability or competitive displacement. 
This raises a key question: how does partner turnover influence host demography and long-term fitness?
We studied the fitness consequences of partner diversity and turnover in the sequential multi-species food-for-protection mutualism involving an EFN-bearing plant (tree cholla cactus \textit{Cylindropuntia imbricata}) and multiple species of nectar-feeding ants (\textit{Crematogaster opuntiae}, \textit{Liometopum apiculatum}, and other less frequent ants) that provide defense from herbivores. 
We used along-term dataset on plant vital rates and partner identity and Bayesian hierarchical statistical models to quantify the demographic consequences of association with different ant partners and the frequency and direction of turnover between mutually exclusive partner associations across the plant life cycle.
We then constructed a stochastic, multi-state Integral Projection Model (IPM) to simulate how different patterns of partner richness, composition, and sequential turnover affect population fitness under fluctuating environmental conditions.
We found that, while ant partners had different impacts on plant vital rates (\textit{C. opuntiae}-tended plants had advantages in growth and survival when small, and \textit{L. apiculatum}-tended plants had floral viability advantages), plant fitness was effectively insensitive to partner identity and number, because effects on the highest-sensitivity vital rates were consistent across partner species. 
Furthermore, there was no evidence that benefits of partner diversity manifest in the context of fluctuating environments (i.e., the portfolio effect). 
In contrast to much of the previous literature, this study highlights that demographic benefits of mutualism can be surprisingly robust to diversity and turnover of partners.

\newpage{}
\section*{Introduction} \label{res:introduction}
%% introduce multispecies mutualisms and sequential mutualisms with definitions
Mutualisms, species interactions where all participants receive net benefits, are key components of ecological communities and can enhance the fitness and population growth rates of the partners involved.  \linelabel{res:27}
While traditionally framed as pairwise (+/+) interactions, mutualisms are often embedded within multispecies networks in which a focal species interacts with multiple partners over space or time \citep{Bascompte2009, Palmer2010}.
In such systems, partner diversity can arise either from simultaneous interactions or sequential ones, where individual hosts experience turnover in partner identity over time \citep{Djieto-Lordon2005, Ness2006}. \linelabel{res:L22}
A focal mutualist may interact with multiple guilds of partner types (e.g., plants that interact with pollinators, seed dispersers, soil microbes, and ant defenders) or with multiple partner species within the same guild (e.g., plants visited by multiple pollinator species). 
Understanding how this kind of sequential partner diversity affects host fitness is a pressing question in mutualism ecology—especially under environmental variability, where different partners may confer benefits unevenly across space or time  \citep{rogers2014, Thibaut2012, Frederickson2005, Palmer2010}. \linelabel{res:L39}

%% Look at mechanisms which show how diversity impacts demography
In this study, we evaluate several mechanisms through which partner diversity might influence mutualistic function, drawing from broader biodiversity–ecosystem function theory \citep{Barrett2015,Ushio2020,Winfree2020}.
First, sampling effect occurs in systems with a distinct hierarchy of fitness effects (a consistent ranking of best to worst mutualists), when diversity increases the likelihood of including a high-quality partner \citep{Frederickson2013,Batstone2018}. \linelabel{res:mechanims_math} \linelabel{res:L45}
There is no benefit of diversity here \emph{per se}, rather, missed opportunity costs arise in systems with sequential partner associations, where less beneficial partners may preclude colonization by better mutualists \citep{Miller2007}.
Second, complementarity occurs in systems where partners differ functionally, when diversity increases the net benefits received by focal mutualist beyond what a single partner species could offer \citep{Batstone2018}.
Positive diversity effects through complementarity need not be additive: interference or synergies between partners can make their combined effect different than the expected from the sum of complementary functions \citep{Afkhami2014}. 
Even within a single mutualist guild, the benefits conferred by alternative partner species can vary by method , efficiency, or timing of benefit conferral \citep{Stachowicz2005,Bronstein2006,Stanton2003}. \linelabel{res:L53}
Third, portfolio effects stabilize host fitness across variable environments when different mutualists respond asynchronously to environmental fluctuations \citep{Batstone2018, Loreau1999, Lazaro2022,Ollerton2006,Alarcon2008}. \linelabel{res:portfolio_asynchronous} 
In this context, the portfolio effect implies that inter-annual variation in partner identity—especially if driven by environmental conditions—could stabilize overall plant fitness across years.
For example, if a high quality defensive partner is more likely to occur during dry years and a moderately defensive partner dominates in wet years, the shifting presence of these species could buffer the focal plant from herbivore damage under variable conditions. \linelabel{res:portfolio_theory}
Critically, portfolio effects require asynchronous environmental responses among partners, such that declines in one partner's benefits are offset by others, stabilizing the focal mutualist's fitness over time \citep{Loreau1999, Tilman1994, Hautier2014}.
This concept is well developed in biodiversity–ecosystem functioning literature, but has rarely been applied to mutualisms in which partner diversity is realized through turnover rather than coexistence.
While portfolio effects have been well studied in plant and animal communities, their role in mutualisms—particularly where partner diversity is sequential (turnover) rather than simultaneous—is less explored \citep{Ushio2020}.

%% Partner turnover
Partner diversity can affect a focal mutualist differently depending on whether partners are present simultaneously or sequentially (i.e., via partner turnover) \citep{Djieto-Lordon2005, Ness2006, Bruna2014, Barrett2015, Ushio2020, Dattilo2014}. 
In simultaneous associations partner interactions frequently lead to competition and interference \citep{Kiers2003,Batstone2018, Tgaard2015, Wulff2008}, reducing the net benefits received by the focal mutualist. \linelabel{res:L63}
Sequential associations have been shown to reduce these competitive partner-partner interactions, thereby boosting the focal mutualists benefits \citep{Wolfin2023,Stachowicz2005_2} and leading to more stable interactions \cite{Johnson2019}.
Turnover can happen at different timescales, from minutes to years \citep{Oliveira1999,Horvitz1986}. 
The timing, frequency, and directionality of partner turnover can influence the net fitness benefits to the host, particularly if turnover allows accumulation of complementary benefits or buffers environmental variability via portfolio effects \citep{Sachs2004,Fiala1994, Ushio2020}. 
If partner identity changes systematically across ontogeny, such as through succession of partners aligned with host life stages, this may further affect fitness outcomes \citep{Fonseca2003,Noe1994,Dejean2008}.
For example, plant susceptibility to enemies can change across life stages \citep{Boege2005,Barton2010}, so the benefits of a diverse guild of defensive mutualists are greatest when more defensive partner species align with more vulnerable life stages \citep{Djieto-Lordon2005,Dejean2008}.
Thus, the duration and order of interactions may be just as important as the identities of the interacting partners.

%% Defensive ant-plant mutualisms
Defensive ant-plant mutualisms -- where plants provide food and/or housing to ants that in turn defend them from enemies -- are widespread interactions that offer valuable model systems for the ecology and evolution of mutualism \citep{Bronstein1998, Bronstein2006}. 
Extrafloral nectar (EFN) -bearing plants can serve as dietary resources that promote ant abundance and colony size \citep{Byk2011, Ness2009, Ness2006, Donald2022}.
In turn, presence of defensive ant partners is often linked to reductions in herbivory  \citep{Trager2010, Rudgers2004} and demographic advantages for the plant partner \citep{Baez2016}.
Defensive ant-plant mutualisms are commonly multi-species, where a guild of ant partner species share, and often compete for, a plant mutualist \citep{Bronstein1998, Beattie1985, Trager2010, Agrawal1998}.
Ant partners can vary in their ability to deter herbivores \citep{Bruna2014}, and visitation by low quality ant partners can prevent visitation by higher quality partners, causing a reduction in fitness through missed opportunity costs \citep{Fraser2001, Frederickson2005}.
Susceptibility to herbivory can also vary significantly throughout the life stages of the plant \citep{Boege2005}, suggesting that the order and timing of successive partners is important to the fitness impacts of the combined partner guild \citep{Barton2010, Boege2005, Fonseca2003}.
Finally, herbivore identity and pressure can vary inter-annually \citep{Wetzel2023}, much like mutualist identity and presence, meaning the threat plants face can vary just as much as the protection they receive due to temporal stochasticity. 
Previous studies have investigated how ant partner diversity affects plant fitness \citep{Palmer2010,Fiala1994,Gaume1998,Dattilo2014,Ludka2015}. 
However, little is known about the combined effects of partner identity, directional partner turnover, and temporal stochasticity, particularly because the long-term data needed to quantify fluctuations in ant-plant interactions are rarely available. 
Specifically, sequential associations require much longer-term data in order to analyze due to an accumulation of partners across time rather than the ability to capture multiple at once. \linelabel{res:long-term-data}

%% What this study set out to do
This study examined the consequences of partner diversity in a food-for-protection mutualism between the tree cholla cactus (\textit{Cylindropuntia imbricata}), a long-lived EFN-bearing plant, and multiple species of ant partners.
Tree cholla are tended by two common ant partners and several additional rarer species, all of which collect EFN in exchange for defense against herbivores.
These ant species locally co-occur but individual plants are typically tended by only one species that patrols the plant around-the-clock and maintains control of the plant's nectar resources, usually for an entire growing season \citep{Ohm2014, Donald2022}. 
Switches between partner species, or between vacancy and ant occupancy, commonly occur from one growing season to the next \citep{Miller2007}. 
Prior experiments suggested a hierarchy of mutualist quality, with \textit{Liometopum apiculatum} providing strong anti-herbivore defense and \textit{Crematogaster opuntiae} having net negative effects because herbivore deterrence is outweighed by deterrence of pollinators \citep{Miller2007,Ohm2014}.  \linelabel{res:L111}
Even the basic quantification of how different ant species affect host demography across the plant life cycle is rarely possible, particularly in long-lived perennials. 
Most studies evaluate short-term fitness proxies like seed set or herbivory at a single life stage \citep{Palmer2010, Bruna2014, Trager2010}, while few trace demographic effects across multiple stages.
Similarly, while turnover of mutualist partners is known to occur \citep{Ness2006, Djieto-Lordon2005}, long-term data documenting patterns and consequences of partner switching are uncommon, especially in field systems. \linelabel{res:long-term-data3}
Thus, this study provides unique empirical insight into how mutualist quality and temporal partner dynamics shape host population trajectories.

%% Models and questions
Here we used a unique long-term data set that allows us to explore mutualistic associations with multiple partner species, longitudinal dynamics of partner turnover, and demographic effects of alternative partners that vary across plant size structure and across nearly 20 years of inter-annual fluctuations. 
The long-term nature of this observational dataset gives us the power to investigate whether and through which mechanism(s) partner diversity affects the fitness benefits of ant visitation. \linelabel{res:long-term-data2}
Specifically, we asked:
	\begin{enumerate}	
		\item{What are the vital rate effects of association with alternative partners and how do these effects fluctuate across years?}
		\item{What are the frequency and direction of partner turnover across the plant life cycle?}	
		\item{What is the net effect of partner diversity on plant fitness, and what mechanism(s) explain(s) this effect?}
	\end{enumerate}
To answer these questions we used a hierarchical Bayesian statistical approach to estimate demographic vital rates for hosts in different states of ant occupancy and to quantify state-dependent partner turnover from the long-term data. 
We then used a stochastic, multi-state integral projection model (IPM) that combines diverse effects on vital rates and pathways of partner turnover to quantify effects of partner diversity on plant fitness.

@@

\subsection*{Data Collection}
This study is based on long-term data spanning 2004 to 2023 that documented plant demographic rates in relation to their ant partner status. 
From 2004 to 2008, we censused 134 plants distributed across three spatial blocks. 
This initial census group was discontinued in 2009, when we established six 30 $\times$ 30-meter plots and tagged all tree cholla within those plots. 
Two additional 30 $\times$ 30-meter plots were added in 2011, and this group of eight plots (mean $\pm$ SD Tree cholla per plot: 87.30 $\pm$ 18.28) has since been censused annually through 2023 (with the exception of 2020 due to the pandemic shutdown). 
These censuses were exhaustive, tracking every single plant and searching the entire plot for any new ones. \linelabel{res:exhaustive}
For all plants, in May or early June of each year we recorded plant survival since the last survey (if the plant had any green flesh left it was considered alive) and, for survivors, we recorded height (cm), maximum crown width (cm), and crown width perpendicular to the maximum (cm) (sections with green were measured for size).
Size measurements were used to calculate plant volume ($cm^3$) based on the volume of an elliptical cone. 
We measured reproduction by counting flowerbuds, and in most years we distinguished between flowerbuds that were viable versus aborted (this differentiation was a visual assessment). 
We recorded the ant species presence and identity (or vacancy if no ants present).
Occurrences of more than one ant species on one plant were rare (less than 5\% of observations), and for the purpose of this analysis we classified the plant as being occupied by the more abundant species. 
Plots were searched for new recruits each year, and these were added to the census.
These data allowed us to link each plant's demographic fate (survival, growth, and reproduction) to its state of ant visitation. 
In total, the data set includes a total of 9,787 observations of 1,141 individuals across 15 complete transition years (spanning May/June of year $t$ to May/June of year $t+1$) \citep{DataCholla}. 
In addition to missing the year 2020, there are gaps in the time series where we switched plots or plants (and thus broke up transition years for growth/survival and partner turnover) or where we did not distinguish between viable and aborted flower buds.

To complete the tree cholla life cycle, we used additional, smaller data sets from previously published studies to estimate seed and seed bank parameters. 
These studies were conducted at the same sites using the same plants as our study. \linelabel{res:data_origins}
Ohm et al. \citeyear{Ohm2014} provide data on the number of seeds per fruit for plants tended by \textit{L. apiculatum}, \textit{C. opuntiae}, or no ants (experimental exclusion), accounting for their effects on pollinator visitation. 
Elderd and Miller \citeyear{elderd2016quantifying} provide data on seed entry to the seed bank and seedling germination and survival rates.

@@

\subsubsection*{Partner diversity simulations}
Using the fully parameterized multi-state IPM, we conducted simulations to quantify how diversity and identity of ant partners influenced plant fitness. 
From the full version of the model (described above) corresponding to the observed assemblage of partners and observed frequencies of partner transition, we created treatments corresponding to all eight ``counter-factual'' scenarios of diversity and composition: no ant partners (complete vacancy); one ant partner (\textit{C. opuntiae} only, \textit{L. apiculatum} only, Other only); two partners (all pairwise combinations of \textit{C. opuntiae}, \textit{L. apiculatum}, and Other); and three partners (observed scenario of all ant states).
These simulations were made possible by extrapolating ant-specific demographic performance across the size distribution, even for combinations of size and ant occupancy that were rarely observed. 
For example, the no-partner scenario modeled a hypothetically ant-free cactus population, even though no such population exists to our knowledge, by applying the statistical knowledge gleaned from vacant plants across the size distribution. 
We refer to stochastic fitness associated with partner number or identity using superscripts, e.g. $\lambda^{0}_{S}$ for vacant plants (zero partners), $\lambda^{1+}_{S}$ for any state of ant tending, $\lambda^{C}_{S}$ for tending by only \textit{C. opuntiae}, $\lambda^{CO}_{S}$ for tending by \textit{C. opuntiae} and Other ants, etc.

In all scenarios that included any ant partners, we preserved the observed pattern of size-dependent vacancy/occupancy (estimated through the ant transition sub-model) and manipulated partner identity conditional on occupancy by any ant. 
This means, for example, that the \textit{C. opuntiae}-only scenario included two possible states, vacancy and occupied by \textit{C. opuntiae}. 
While our statistical models allow us to extrapolate the demographic performance of ant-tended plants to small sizes that are typically vacant, the natural history of this system tells us that this is not biologically sensible. 
Small, non-reproductive plants are typically vacant because they do not produce extrafloral nectar, and once plants begin producing nectar they are nearly always ant-tended \citep{Miller2014}. 
Our simulations preserved this basic biology, avoiding tiny ant-occupied plants that do not and could not occur in nature.

@@

\begin{figure}[H]
	\includegraphics[width = 0.75\linewidth]{Figures/grow_contour_v2.png}
	\caption{The next predicted size of tree cholla based on previous size in relation to ant partner (a, \textit{C. opuntiae}; b, \textit{L. apiculatum}; c, Other; d, Vacant, e, all states combined). Points show observed size transitions, colored lines are the mean next size, and contours show probability density of the skewed normal growth model.  
	Dashed lines indicate extrapolation while solid lines indicate the range of observed data.
	They gray dashed line indicates no growth (areas above this line show growth while areas under it show shrinkage).}
	\label{fig:Grow}
\end{figure}

@@

\begin{figure}[H]
	\includegraphics[width=0.95\linewidth]{Figures/transition.png}
	\caption{The probability of ant state based on plant size and previous ant state. Each panel shows size-dependent probabilities of the next ant state conditional on the previous ant state and size. Solid lines represent predictions of the multinomial statistical model and points show data binned over size intervals, where point size is proportional to the number of observations.}
	\label{fig:Ant_Transition}
\end{figure}
